|

### 2023-02-04 paper.Rmd
Sat Feb  4 22:00:06 CET 2023: update-remote.sh

---
title: |
  <center>Pre-Trained Denoising Autoencoders Long Short-Term Memory Networks as probabilistic Models for Estimation of Distribution Genetic Programming</center>
  <center>--------------------------------------------------------</center>
  <center>Student: Roman Höhn</center>
  <center>Date of Birth: 1991-04-14</center>
  <center>Place of Birth: Wiesbaden, Hesse</center>
  <center>Student ID: 2712497</center>
  <center>Supervisor: David Wittenberg</center>
  <center>--------------------------------------------------------</center>
  <center>Master Thesis - M.Sc. Business Education</center>
  <center>FB 03: Chair of Business Administration and Computer Science</center>
  <center>Johannes Gutenberg University Mainz</center>
  |
date: "Date of Submission: `r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: true
    extra_dependencies: ["float"]
    fig_caption: yes
    number_sections: true
    toc_depth: 3
    includes:
      in_header: [header_files/settings.tex]
indent: false
fontsize: 11pt
bibliography: ./ref/ref.bib
link-citations: true
fig.width: 4in
fig.height: 3in
csl: csl/association-for-computing-machinery.csl
urlcolor: blue
linkcolor: blue
toccolor: blue
#csl: csl/harvard-cite-them-right.csl
#csl: csl/springer-science-reviews.csl
include-before: '`\thispagestyle{empty} \newpage{}`{=latex}'
---

@@

```{r airfoil-fullRun-2hl-maxIndSize-params, echo=FALSE, message=FALSE, warning=FALSE}
fig.pos="H"
knitr::kable(
  read.csv("./tables/airfoil_fullRun_2hl_maxIndSize_params", sep = "?"),
  caption = "DAE-GP - Hyperparameter"
)%>% row_spec(0,bold=TRUE )
```

@@

Another interesting metric to examine in population-based optimization algorithms is the diversity of the current generations population. Since individual solutions in DAE-GP are computer programs in the form of parse trees, I selected the normalized Levenshtein edit distance as my primary metric for tracking population diversity. Figure \@ref(fig:airfoil-popDiversity) shows the median population diversity for both DAE-GP algorithms and regular GP as a benchmark. While the population diversity steadily decreases over the generations for standard GP, DAE-GP variants strongly increase the population diversity in the first generations and keep them at a higher level during the evolutionary search. Here it can be observed, that pre-training shows a positive effect by increasing population diversity in comparison to regular DAE-GP (Median over all generations for DAE-GP: `r round(REG_MED_LEVDIV, 3)`, Pre-Trained DAE-GP: `r round(PT_MED_LEVDIV, 3)`, Standard-GP `r round(GP_MED_LEVDIV, 3)`)

@@

To gather more confidence in the previous results, the pre-training strategy for DAE-GP was applied for three additional real world symbolic regression problems and tested for statistically significant differences in the underlying distributions. Table \@ref(tab:full-run-realWorldSymReg-problems) briefly summarizes the benchmark problems.

```{r full-run-realWorldSymReg-problems, echo=FALSE, message=FALSE, warning=FALSE}
fig.pos="H"
knitr::kable(
  read.csv("./tables/full_run_realWorldSymReg_problems.csv"),
  caption="Real World Symbolic Regression Benchmark Problems",
  format = "markdown"
)%>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")
```
To test for statistical differences between the distributions of results achieved by either pre-trained or regular DAE-GP, I selected the  
nonparametric Mann-Whitney-U Test ([@mwu1], [@mwu2]) to test if the distribution .... Additionaly the effect strength of pre-training is computed by using Cliffs Delta[@Cliff1993DominanceSO].

## Fitness

@@

```{r symReg-Fitness, echo=FALSE, fig.cap="Fitness over 30 Generations - Summary - Real World Symbolic Regression", out.width="80%"}
knitr::include_graphics("~/masterThesis/master_thesis/img/symbolicRegressionSummarized/median_fitness_byGens.png")
```

Table \@ref(tab:full-run-realWorldSymReg-fitness) summarizes the results for the median best fitness achieved after 30 generations and shows the results of computing the statistical tests. Higher fitness values (lower RMSE) are printed bold and p-values are marked with asterisk (1,2,3) for $\alpha$ levels ($0.1$, $0.05$, $0.01$).

@@

knitr::kable(
 df_bold,
 format = "markdown",
 col.names = col_names,
 align = "l",
 caption = "Median Best Fitness after 30 generations - Summary - Real World Symbolic Regression"
) %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")

@@

## Solution Size

@@

```{r symReg-Size, echo=FALSE, fig.cap="Size of the best Solution over 30 Generations - Summary - Real World Symbolic Regression", out.width="80%"}
knitr::include_graphics("~/masterThesis/master_thesis/img/symbolicRegressionSummarized/median_bestSol_size_byGens.png")
```

@@

knitr::kable(
 df_bold,
 format = "markdown",
 col.names = col_names,
 align = "l",
 caption = "Median Size of the best solution after 30 generations - Summary - Real World Symbolic Regression"
) %>% row_spec(0,bold=FALSE) %>% kableExtra::kable_styling(latex_options = "hold_position", font_size = 7)

```
## Population Diversity

![Median Population Diversity over Generations](~/masterThesis/master_thesis/img/symbolicRegressionSummarized/median_levDistance_byGen.png){height=80%}

Testing for statistical differences in the underlying distributions yields the P-Values listed in table X.

@@

knitr::kable(
 df_bold,
 format = "markdown",
 col.names = col_names,
 align = "l",
 caption = "Median Population Diversity over Generations - Symbolic Regression"
) %>% 
  kable_styling(latex_options = "scale_down", font_size = 7)

@@

## Training epochs per Generation

![Median Number of Training Epochs per Generation - Symbolic Regression](~/masterThesis/master_thesis/img/symbolicRegressionSummarized/median_epochsTrained.png){height=80%}

@@

knitr::kable(
 df_bold,
 format = "markdown",
 col.names = col_names,
 align = "c",
 caption = "Median Number of Training Epochs per Generation - Symbolic Regression"
) %>% 
  kable_styling(latex_options = "scale_down", font_size = 7)

>>>>

---
title: |
  <center>Pre-Trained Denoising Autoencoders Long Short-Term Memory Networks as probabilistic Models for Estimation of Distribution Genetic Programming</center>
  <center>--------------------------------------------------------</center>
  <center>Student: Roman Höhn</center>
  <center>Date of Birth: 1991-04-14</center>
  <center>Place of Birth: Wiesbaden, Hesse</center>
  <center>Student ID: 2712497</center>
  <center>Supervisor: David Wittenberg</center>
  <center>--------------------------------------------------------</center>
  <center>Master Thesis - M.Sc. Business Education</center>
  <center>FB 03: Chair of Business Administration and Computer Science</center>
  <center>Johannes Gutenberg University Mainz</center>
  |
date: "Date of Submission: `r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: true
    extra_dependencies: ["float"]
    fig_caption: yes
    number_sections: true
    toc_depth: 3
    includes:
      in_header: [header_files/settings.tex]
indent: false
fontsize: 12pt
bibliography: ./ref/ref.bib
link-citations: true
fig.width: 4in
fig.height: 3in
csl: csl/association-for-computing-machinery.csl
# csl: csl/harvard-cite-them-right.csl
# csl: csl/springer-science-reviews.csl
urlcolor: blue
linkcolor: blue
toccolor: blue
github-repo: https://github.com/roman91DE/master_thesis
include-before: '`\thispagestyle{empty} \newpage{}`{=latex}'
---

@@

```{r airfoil-fullRun-2hl-maxIndSize-params, echo=FALSE, message=FALSE, warning=FALSE}
fig.pos="H"
knitr::kable(
  read.csv("./tables/airfoil_fullRun_2hl_maxIndSize_params", sep = "?"),
  caption = "DAE-GP - Hyperparameter",
  format = "markdown"
)%>% row_spec(0,bold=TRUE )
```

@@

Another interesting metric to examine in population-based optimization algorithms is the diversity of the current generations population. Since individual solutions in DAE-GP are computer programs in the form of parse trees, I selected the normalized Levenshtein edit distance [@NormLevDistance] as my primary metric for tracking population diversity. Figure \@ref(fig:airfoil-popDiversity) shows the median population diversity for both DAE-GP algorithms and regular GP as a benchmark. While the population diversity steadily decreases over the generations for standard GP, DAE-GP variants strongly increase the population diversity in the first generations and keep them at a higher level during the evolutionary search. Here it can be observed, that pre-training shows a positive effect by increasing population diversity in comparison to regular DAE-GP (Median over all generations for DAE-GP: `r round(REG_MED_LEVDIV, 3)`, Pre-Trained DAE-GP: `r round(PT_MED_LEVDIV, 3)`, Standard-GP `r round(GP_MED_LEVDIV, 3)`)

@@

To gather more confidence in the previous results, I applied the pre-training strategy for DAE-GP to three additional real world symbolic regression problems. Table \@ref(tab:full-run-realWorldSymReg-problems) briefly summarizes the benchmark problems.

```{r full-run-realWorldSymReg-problems, echo=FALSE, message=FALSE, warning=FALSE}
fig.pos="H"
knitr::kable(
  read.csv("./tables/full_run_realWorldSymReg_problems.csv"),
  caption="Real World Symbolic Regression Benchmark Problems",
  format = "markdown"
)%>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")
```
To test for statistical differences of the results achieved by either pre-trained or regular DAE-GP, I selected the nonparametric Mann-Whitney-U Test ([@mwu1], [@mwu2]) to test for differences in the distribution underlying the samples sourced from my experiments. Additionaly, I included the effect strength of pre-training as measured by Cliffs Delta [@Cliff1993DominanceSO] into the statistical analysis. Better median values (e.g. higher fitness) are printed bold and p-values are marked with asterisk (1,2,3) for $\alpha$ levels ($0.1$, $0.05$, $0.01$).

Note that the results for the real world symbolic regression problems inside this chapter, except for the airfoil problem (see subsection [Influence on Run-Time]), have not been analyzed for their total run-time. This is due to the fact, that experiments were run on different computers architectures which resulted in strongly biased results.

## Influence on Fitness

@@

```{r symReg-Fitness, echo=FALSE, fig.cap="Fitness over 30 Generations - Real World Symbolic Regression", out.width="90%"}
knitr::include_graphics("~/masterThesis/master_thesis/img/symbolicRegressionSummarized/median_fitness_byGens.png")
```

Table \@ref(tab:full-run-realWorldSymReg-fitness) summarizes the results for the median best fitness achieved after 30 generations and shows the results of computing the statistical tests.

@@

knitr::kable(
 df_bold,
 format = "markdown",
 col.names = col_names,
 align = "l",
 caption = "Median Best Fitness after 30 generations - Real World Symbolic Regression"
) %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")

@@

## Influence on Solution Size

@@

```{r symReg-Size, echo=FALSE, fig.cap="Size of the best Solution over 30 Generations - Real World Symbolic Regression", out.width="90%"}
knitr::include_graphics("~/masterThesis/master_thesis/img/symbolicRegressionSummarized/median_bestSol_size_byGens.png")
```

@@

knitr::kable(
 df_bold,
 format = "markdown",
 col.names = col_names,
 align = "l",
 caption = "Median Size of the best solution after 30 generations - Real World Symbolic Regression"
) %>% row_spec(0,bold=FALSE) %>% kableExtra::kable_styling(latex_options = "hold_position", font_size = 7)

```
## Influence on Population Diversity

Figure \@ref(fig:symReg-LevDiv) shows the median levenshtein edit distance as a metric for population diversity over the evolutionary search.

```{r symReg-LevDiv, echo=FALSE, fig.cap="Population Diversity over 30 Generations - Real World Symbolic Regression", out.width="90%"}
knitr::include_graphics("~/masterThesis/master_thesis/img/symbolicRegressionSummarized/median_levDistance_byGen.png")
```

The results of the statistical analysis is summarized in table \@ref(tab:full-run-realWorldSymReg-popDiversity).

@@

for(i in 1:nrow(df)){
  if(df[i, "DAE.GP"] > df[i, "Pre.Trained_DAE.GP"]){
    df_bold[i, "DAE.GP"] <- sub("^(.*)$", "**\\1**", as.character(round(df[i, "DAE.GP"],2)))
    df_bold[i, "Pre.Trained_DAE.GP"] <- as.character(round(df[i, "Pre.Trained_DAE.GP"],2))
    }
  else if (df[i, "DAE.GP"] < df[i, "Pre.Trained_DAE.GP"]) {
    df_bold[i, "Pre.Trained_DAE.GP"] <- sub("^(.*)$", "**\\1**", as.character(round(df[i, "Pre.Trained_DAE.GP"],2)))
  df_bold[i, "DAE.GP"] <- as.character(round(df[i, "DAE.GP"],2))
  }
}
# shorten column names
col_names <- c("Problem", "Hid.Layers", "DAE-GP", "Pre-Trained", "P-Value", "Cliffs-Delta")

knitr::kable(
 df_bold,
 format = "markdown",
 col.names = col_names,
 align = "l",
 caption = "Median Population Diversity over 30 Generations - Symbolic Regression"
) %>% 
  kable_styling(latex_options = "scale_down", font_size = 7)

@@

## Influence on the number of Training epochs per Generation

Figure \@ref(fig:symReg-EpochsPerGen) shows the median number of epochs spend on training each generations DAE-LSTM network until the termination criterion was satisfied.

```{r symReg-EpochsPerGen, echo=FALSE, fig.cap="Training Epochs over 30 Generations - Real World Symbolic Regression", out.width="90%"}
knitr::include_graphics("~/masterThesis/master_thesis/img/symbolicRegressionSummarized/median_epochsTrained.png")
```

The results of the statistical analysis is summarized in table \@ref(tab:full-run-realWorldSymReg-epochsPerGen).

@@

knitr::kable(
 df_bold,
 format = "markdown",
 col.names = col_names,
 align = "l",
 caption = "Median Number of Training Epochs per Generation - Symbolic Regression"
) %>% 
  kable_styling(latex_options = "scale_down", font_size = 7)
